---
title: "HGEN 486 Final Project: The Gibbs SLAM-pler"
author: Austin Szatrowski
monofont: iA Writer Mono S
output:
  html_document
---
```{r}
library(tidyverse)
```

## Simulate Data
For now, we stick with aligned reads that have already been masked versus reference:

* we're constraining such that all allele frequencies are equal (should be true; even errors and end effects shouldn't affect underlying substitution rates... unless there are within-gene transcription effect? who knows)

ALSO: this is all per-gene at the moment. we'll need a scale-up for genome-wide and to show that it works for a range of read counts and new RNA % levels.

```{r}
sim_reads <- function(read_count, sub_rates_matrix, total_new_rna_prop) {
  num_sites <- ncol(sub_rates_matrix)
  labels <- sample(
    c("old", "new"),
    prob = c(1 - total_new_rna_prop, total_new_rna_prop),
    size = read_count,
    replace = TRUE
  )
  reads <- matrix(0, read_count, num_sites)
  for (i in seq_len(read_count)) {
    reads[i, ] <- rbinom(num_sites, rep(1, num_sites), sub_rates_matrix[labels[i], ])
  }
  return(list(reads = reads, labels = labels))
}

sub_rate_matrix <- function(sub_rate_old, sub_rate_new, read_length) {
  # given T>C substitution rates in old and new RNA, build a 2 x 150bp matrix of per-site rates
  sub_rates <- matrix(
    c(
      rep(sub_rate_old, read_length),
      rep(sub_rate_new, read_length)
    ),
    nrow = 2,
    byrow = TRUE
  )
  rownames(sub_rates) <- c("old", "new")
  return(sub_rates)
}
rate_matrix <- sub_rate_matrix(
  sub_rate_old = 0.01,
  sub_rate_new = 0.2,
  read_length = 150
)
sim <- sim_reads(
  read_count = 500,
  sub_rates_matrix = rate_matrix,
  total_new_rna_prop = 0.2
)
```

```{r}
normalize <- function(x) {
  return(x / sum(x))
}
loglik_read_rates <- function(read, f) {
  # read: vector of length J (sites), binary 0/1
  # f: 2 x J matrix (row 1 = old rates, row 2 = new rates)
  # returns: vector of length 2 (loglik of this read under old, under new)
  # for each component (old/new), calculate log P(read | rates)
  # loglik = sum over sites of: x_j * log(f) + (1-x_j) * log(1-f)
  loglik <- rep(NA, 2)
  for (k in 1:2) {
    loglik[k] <- sum(read * log(f[k, ]) + (1 - read) * log(1 - f[k, ]))
  }
  return(loglik)
}

sample_z <- function(reads, f, pi_new) {
  # reads: N x J matrix of reads (rows = reads, cols = sites)
  # f: 2 x J matrix of conversion rates (row 1 = old, row 2 = new)
  # pi_new: scalar, current proportion of new RNA
  # returns: vector of length N with assignments (1 = old, 2 = new)
  N <- nrow(reads)
  # calculate log-likelihood of each read under each component
  loglik_matrix <- apply(
    X = reads,
    MARGIN = 1,  # iterate over reads
    FUN = loglik_read_rates,
    f = f
  )
  # loglik_matrix is now 2 x N (row 1 = loglik under old, row 2 = loglik under new)
  # add log prior (mixing proportions)
  log_pi <- log(c(1 - pi_new, pi_new))
  loglik_matrix <- loglik_matrix + log_pi  # broadcasting across columns
  # convert to probabilities (exponentiate and normalize)
  lik_matrix <- exp(loglik_matrix)
  p_z_given_read <- apply(
    X = lik_matrix,
    MARGIN = 2,  # normalize each column (each read)
    FUN = normalize
  )
  # p_z_given_read is now 2 x N, with columns summing to 1
  # sample assignments
  z <- numeric(N)
  for (i in 1:N) {
    z[i] <- sample(
      x = 1:2,
      size = 1,
      prob = p_z_given_read[, i]
    )
  }
  return(z)
}

sample_f <- function(reads, z) {
  J <- ncol(reads) # number of sites
  MatrixF <- matrix(NA, nrow = 2, ncol = J) # allele freq matrix, 2 x J
  for (i in seq_len(2)) {
    sample_size <- sum(z == i)
    if (sample_size == 0)
      number_of_ones <- rep(0, J)
    else
      number_of_ones <- colSums(reads[z == i, ])
    MatrixF[i, ] <- rbeta(J, number_of_ones + 1, sample_size - number_of_ones + 1)
  }
  return(MatrixF)
}

sample_pi <- function(z, prior_alpha = 1, prior_beta = 5) {
  # estimate pi from z.
  # if prior_beta >> prior_alpha, put a stronger prior on new RNA being rare
  n_new <- sum(z == 2)
  n_old <- sum(z == 1)
  pi_new <- rbeta(1, n_new + prior_alpha, n_old + prior_beta)
  return(pi_new)
}
```
```{r def sampler}
gibbs <- function(read_data, niter = 100) {
  n <- nrow(read_data)
  J <- ncol(read_data)
  pi_init <- c(0.1, 0.9)

  z_out <- matrix(NA, nrow = niter, ncol = n)
  pi_out <- rep(NA, niter)
  f_out <- array(0, dim = c(niter, 2, J))

  z <- sample(x = 2, size = n, replace = TRUE, prob = pi_init)
  z_out[1, ] <- z

  f <- sample_f(read_data, z)
  f_out[1, , ] <- f

  pi_new <- sample_pi(z)
  pi_out[1] <- pi_new

  for (i in 2:niter) {
    f <- sample_f(read_data, z)
    z <- sample_z(read_data, f, pi_new)
    pi_new <- sample_pi(z, prior_alpha = 1, prior_beta = 5)
    z_out[i, ] <- z
    f_out[i, , ] <- f
    pi_out[i] <- pi_new
  }

  return(list(z = z_out, f = f_out, pi = pi_out))
}
```
```{r run sampler}
run <- gibbs(
  read_data = sim$reads
)

image(run$z)
image(run$f[ , 1, ])
plot(run$pi, type = "l")
print(mean(run$pi[50:100]))
```
